
# Beta diversity

Beta diversity is another name for sample dissimilarity. It quantifies
differences in the overall taxonomic composition between two samples.

Common indices include Bray-Curtis, Unifrac, Jaccard index, and the
Aitchison distance. Each of these (dis)similarity measures emphasizes
different aspects. For example, UniFrac incorporates phylogenetic
information, and Jaccard index ignores exact abundances and considers
only presence/absence values. For more background information
and examples, you can check the dedicated section in [online
book](https://microbiome.github.io/OMA/microbiome-diversity.html#beta-diversity).


## Examples of PCoA with different settings

Beta diversity estimation generates a (dis)similarity matrix that
contains for each sample (rows) the dissimilarity to any other sample
(columns).

This complex set of pairwise relations can be visualized in
informative ways, and even coupled with other explanatory
variables. As a first step, we compress the information to a lower
dimensionality, or fewer principal components, and then visualize
sample similarity based on that using ordination techniques, such as
Principal Coordinate Analysis (PCoA). PCoA is a non-linear dimension
reduction technique, and with Euclidean distances it is is identical
to the linear PCA (except for potential scaling).

We typically retain just the two (or three) most informative top
components, and ignore the other information. Each sample has a score
on each of these components, and each component measures the variation
across a set of correlated taxa. The top components are then easily
visualized on a two (or three) dimensional display.

Let us next look at some concrete examples.


### PCoA for ASV-level data with Bray-Curtis

Let us start with PCoA based on a Bray-Curtis dissimilarity matrix
calculated at Genus level abundances.


```{r pcoa_asv_bc, message=FALSE}
# Pick the relative abundance table
rel_abund_assay <- assays(tse)$relabundance

# Calculates Bray-Curtis distances between samples. Because taxa is in
# columns, it is used to compare different samples. We transpose the
# assay to get taxa to columns
bray_curtis_dist <- vegan::vegdist(t(rel_abund_assay), method = "bray")

# PCoA
bray_curtis_pcoa <- ecodist::pco(bray_curtis_dist)

# All components could be found here: 
# bray_curtis_pcoa$vectors
# But we only need the first two to demonstrate what we can do:
bray_curtis_pcoa_df <- data.frame(pcoa1 = bray_curtis_pcoa$vectors[,1], 
                                  pcoa2 = bray_curtis_pcoa$vectors[,2])

# Create a plot
bray_curtis_plot <- ggplot(data = bray_curtis_pcoa_df, aes(x=pcoa1, y=pcoa2)) +
  geom_point() +
  labs(x = "PC1",
       y = "PC2", 
       title = "Bray-Curtis PCoA") +
  theme(title = element_text(size = 10)) # makes titles smaller

bray_curtis_plot
```



### PCoA for ASV-level data with Aitchison distance

Now the same using Aitchison distance. This metric corresponds to
Euclidean distances between CLR transformed sample abundance vectors.

```{r pcoa_asv_aitchison}
# Does clr transformation. Pseudocount is added, because data contains zeros. 
tse <- transformCounts(tse, method = "clr", pseudocount = 1)

# Gets clr table
clr_assay <- assays(tse)$clr

# Transposes it to get taxa to columns
clr_assay <- t(clr_assay)

# Calculates Euclidean distances between samples. Because taxa is in columns,
# it is used to compare different samples.
euclidean_dist <- vegan::vegdist(clr_assay, method = "euclidean")

# Does principal coordinate analysis
euclidean_pcoa <- ecodist::pco(euclidean_dist)

# Creates a data frame from principal coordinates
euclidean_pcoa_df <- data.frame(pcoa1 = euclidean_pcoa$vectors[,1], 
                                pcoa2 = euclidean_pcoa$vectors[,2])

# Creates the plot
euclidean_plot <- ggplot(data = euclidean_pcoa_df, aes(x=pcoa1, y=pcoa2)) +
  geom_point() +
  labs(x = "PC1",
       y = "PC22",
       title = "Euclidean PCoA with CLR transformation") +
  theme(title = element_text(size = 12)) # makes titles smaller

euclidean_plot
```


### PCoA aggregated to Phylum level

We use again the Aitchison distances in this example but this time applied to the phylum level.

```{r pcoa_phylum_aitchison}
# Does clr transformation. Psuedocount is added, because data contains zeros. 
tse_phylum <- transformCounts(tse_phylum, method = "clr", pseudocount = 1)

# Gets clr table
clr_phylum_assay <- assays(tse_phylum)$clr

# Transposes it to get taxa to columns
clr_phylum_assay <- t(clr_phylum_assay)

# Calculates Euclidean distances between samples. Because taxa is in columns,
# it is used to compare different samples.
euclidean_phylum_dist <- vegan::vegdist(clr_assay, method = "euclidean")

# Does principal coordinate analysis
euclidean_phylum_pcoa <- ecodist::pco(euclidean_phylum_dist)

# Creates a data frame from principal coordinates
euclidean_phylum_pcoa_df <- data.frame(
  pcoa1 = euclidean_phylum_pcoa$vectors[,1], 
  pcoa2 = euclidean_phylum_pcoa$vectors[,2])

# Creates a plot
euclidean_phylum_plot <- ggplot(data = euclidean_phylum_pcoa_df,
  aes(x=pcoa1, y=pcoa2)) +
  geom_point() +
  labs(x = "PC1",
       y = "PC2",
       title = "Aitchison distances at Phylum level") +  
  theme(title = element_text(size = 12)) # makes titles smaller

euclidean_phylum_plot
```



## Highlighting external variables 

We can map other variables on the same plot for example by coloring
the points accordingly.


### Discrete grouping variable shown with colors


```{r pcoa_genus}
# Adds the variable we later use for coloring to the data frame
euclidean_patient_status_pcoa_df <- cbind(euclidean_pcoa_df,
                             patient_status = colData(tse)$patient_status)

# Creates a plot
euclidean_patient_status_plot <- ggplot(data = euclidean_patient_status_pcoa_df, 
                                        aes(x=pcoa1, y=pcoa2,
                                            color = patient_status)) +
  geom_point() +
  labs(x = "PC1",
       y = "PC2",
       title = "PCoA with Aitchison distances") +
  theme(title = element_text(size = 12)) # makes titles smaller

euclidean_patient_status_plot
```



### PCoA plot with continuous variable

We can do the same as above using any continuous variable. E.g. let us
see how the plotted samples differ in their alpha diversity (as an
example of a continuous variable):


```{r pcoa_coloring}
# Adds coloring information to the data frame, creates new column
euclidean_shannon_pcoa_df <- cbind(euclidean_pcoa_df,
                             shannon = colData(tse)$Shannon_diversity)

# Creates a plot
euclidean_shannon_plot <- ggplot(data = euclidean_shannon_pcoa_df, 
                                 aes(x=pcoa1, y=pcoa2,
                                     color = shannon)) + 
  geom_point() +
  labs(x = "PC1",
       y = "PC2",
       title = "PCoA with Aitchison distances") +
  theme(title = element_text(size = 12)) # makes titles smaller

euclidean_shannon_plot
```

## Estimating associations with an external variable

Next to visualizing whether any variable is associated with
differences between samples, we can also quantify the strength of the
association between community composition (beta diversity) and
external factors.

The standard way to do this is to perform a so-called permutational
multivariate analysis of variance (PERMANOVA). This method takes as
input the abundance table, which measure of distance you want to base
the test on and a formula that tells the model how you think the
variables are associated with each other. 

```{r permanova}
# First we get the relative abundance table
rel_abund_assay <- assays(tse)$relabundance

# again transpose it to get taxa to columns
rel_abund_assay <- t(rel_abund_assay)

# then we can perform the method
permanova_cohort <- vegan::adonis(rel_abund_assay ~ cohort,
                                  data = colData(tse),
                                  permutations = 9999)

# we can obtain a the p value for our predictor:
print(paste0("Different different cohorts and variance of abundance 
              between samples, p-value: ", 
              as.data.frame(permanova_cohort$aov.tab)["cohort", "Pr(>F)"]))
```  

The cohort variable is not significantly associated with
microbiota composition (p-value is over 0.05).

We can, however, visualize those taxa whose abundances drive the
differences between cohorts. We first need to extract the model
coefficients of taxa:


```{r permanova_coefs}
# Gets the coefficients
coef <- coefficients(permanova_cohort)["cohort1",]

# Gets the highest coefficients
top.coef <- sort(head(coef[rev(order(abs(coef)))],20))

# Plots the coefficients
top_taxa_coeffient_plot <- ggplot(data.frame(x = top.coef,
                                             y = factor(names(top.coef),
					     unique(names(top.coef)))),
                                  aes(x = x, y = y)) +
  geom_bar(stat="identity") +
  labs(x="", y="", title="Top Taxa") +
  theme_bw()

top_taxa_coeffient_plot
```

The above plot shows taxa as code names, and it is hard to tell which
bacterial groups they represent. However, it is easy to add human readable
names. We can fetch those from our rowData. Here we use Genus level names:

```{r}
# Gets corresponding Genus level names and stores them to top.coef
names <- rowData(tse)[names(top.coef), ][,"Genus"]

# Adds new labels to the plot
top_taxa_coeffient_plot <- top_taxa_coeffient_plot +
  scale_y_discrete(labels = names) # Adds new labels
top_taxa_coeffient_plot
```


There are many alternative and complementary methods for analysing
community composition. For more examples, see a dedicated section on
beta diversity in the [online
book](https://microbiome.github.io/OMA/).

## Community typing


A dedicated section presenting examples on community typing is in the
[online book](https://microbiome.github.io/OMA/).



## Exercises

 * Visualize community variation with PCoA.

 * Investigate the influence of the data transformations on
   statistical analysis: Visualize community variation with PCoA with
   the following options: 1) Bray-Curtis distances for compositional
   data; 2) Euclidean distances for CLR-transformed data.

 * Community-level comparisons: Use PERMANOVA to investigate whether
   the community composition differs between two groups of individuals
   (e.g. males and females, or some other grouping of your
   choice). You can also include covariates such as age and gender,
   and see how this affects the results.

 * Example [Solutions](08-5-ex-sol-ADHD.html)
