---
title: "Differential abundance analysis"
output: html_notebook
---

## TODO
- Add information about DESeq2
- Polish code
- Polish text

#################################################

In this notebook, we analyse abundances with two different methods: t-test, and DESeq2.
Both of these tests test, if there are statistical significant difference between 
groups.

T-test is  parametric, so it requires normally distributed data. 
It is always wise to check if the data is normally distributed. If it is not then
non-parametric equivalents (t-test = Mann–Whitney U test should be used, if 
normalization is not possible.

DESeq2 is a package for gene expression analysis. It uses binomial dsitributioin to
detect expression changes between groups. In addition to gene analysis, it can also
be used to other purposes.

- Make CLR transformation for the data, then run t-test for every ASV (in a for loop for instance; see t.test() ) between two groups (pick any grouping variable in the metadata, just for demonstration; we can change that later easily if needed) and pick p values t.test(...)$p.value; then do multiple testing correction with p.adjust(), then plot histogram of p-values and adjusted p-values; you can check [this code](https://github.com/microbiome/microbiome/blob/master/inst/extdata/check_wilcoxon.R) for some ideas

Let's first collect the data that we are going to use in analysis.
```{r}
# Does clr transformation. Pseudocount is added, because data contains zeros, and
# clr transformation includes log transformation.
tse <- transformCounts(tse, method = "clr", pseudocount = 1)

# Does transpose, so samples are in rows, then creates a data frame.
abundance_analysis_data <- data.frame(t(assay(tse, "clr")))

# Then we need variable for grouping samples. "Diet" column includes diet information.
# There two groups "western" and "control". Let's include that to the data frame.
abundance_analysis_data <- cbind(abundance_analysis_data, Groups = colData(tse)$Diet)

# Subsets data if, more than 2 different groups
# abundance_analysis_data_sub <- abundance_analysis_data[abundance_analysis_data$Groups == "KO_t1_western" |
#                                                          abundance_analysis_data$Groups == "WT_t2_western",]

```

Next, let's check the normality of data. It can be done with Shapiro-Wilk test.
As we can see, data is not normally distributed.

I don't know if normality test is necessary.##################################

```{r}
# Does Shapiro-Wilk test. Does it only for columns that contain abundances, not for
# column that contain Groups.

normality_test_p <- c()

for (column in 
     abundance_analysis_data[, !names(abundance_analysis_data) %in% "Groups"]){
  # Does Shapiro-Wilk test
  result <- shapiro.test(column)
  
  # Stores p-value to vector
  normality_test_p <- c(normality_test_p, result$p.value)
}

print(paste0("P-values over 0.05: ", sum(normality_test_p>0.05), "/", sum(normality_test_p<=0.05)))
```
Now we can do Student's t-test.
```{r}
# Does t-test. Does it only for columns that contain abundances, not for
# column that contain Groups.

t_test_p <- c()

for (column in 
     names(abundance_analysis_data[, !names(abundance_analysis_data) %in% "Groups"])){
  #print(column)
  result <- t.test(abundance_analysis_data[,column] ~ Groups, data = abundance_analysis_data)
   
   # Stores p-value to vector
   t_test_p  <- c(t_test_p , result$p.value)
}

t_test_p <- data.frame(taxa = 
                         names(abundance_analysis_data[, !names(abundance_analysis_data) 
                                                       %in% "Groups"]),
                       p_raw = t_test_p)

```

Because of multiple tests were made, we need to adjust p-values. That is because, 
these individual tests are not independent. Because of that, if enough tests are made
statistically we will get positive result, even though positive result is extremely 
rare, lets say, e.g., 1 %. 

To avoid false positive results, we must adjust p-values, when multiply tests are made
from same data. With adjustment, the significance levels of individual tests are stricter. 
Here we use Holm method, but there are several other methods also.

```{r}
# Does the adjustment
t_test_p$p_holm <- p.adjust(t_test_p$p_raw, method = "holm")
```

```{r}

df = data.frame(x = c(t_test_p$p_raw, t_test_p$p_holm), 
               type=rep(c("raw", "holm"), c(length(t_test_p$p_raw), length(t_test_p$p_holm))))

t_test_plot <- ggplot(d) + geom_histogram(aes(x=x, fill=type)) +
  xlab("p-value")

t_test_plot
```



- Then add, for the same data, [DESeq2](https://github.com/microbiome/tutorials/blob/master/deseq2.Rmd) example with our present data (exclude ANOVA from that example code)

```{r, echo=FALSE}
library(DESeq2)
library(dplyr)
library(knitr)

# Running the DESeq2 analysis
ds2 <- DESeq2::DESeqDataSet(tse, ~Diet)
dds <- DESeq(ds2)
res <- results(dds)
df <- as.data.frame(res)
df$taxon <- rownames(df)
df <- df %>% arrange(log2FoldChange, padj)

print(head(kable((df))))

```

- Add scatterplot that compares the p-values from both methods

```{r}
mf <- data.frame(df$padj, t_test_p$p_holm)
p <- ggplot(mf, aes(x = (df$padj), y = (t_test_p$p_holm))) +
       labs(x = 'DESeq2 adjusted p-value', y = 't-test adjusted p-value') +
       geom_count() #+ geom_point()
print(p)
```

