---
title: "Differential abundance analysis"
output: html_notebook
---

In this notebook, we analyse abundances with two different methods: t-test, and ANOVA
(Analysis of variance).

Both of these tests test, if there are statistical significant difference between 
groups. However, in t-test, two groups are tested, and in ANOVA all groups are tested
at the same time. 

Both of these methods are parametric, so they require normally distributed data. 
It is always wise to check if the data is normally distributed. If it is not then
non-parametric equivalents (t-test = Mann–Whitney U test and ANOVA = Kruskal-Wallis 
H test) should be used.

- Make CLR transformation for the data, then run t-test for every ASV (in a for loop for instance; see t.test() ) between two groups (pick any grouping variable in the metadata, just for demonstration; we can change that later easily if needed) and pick p values t.test(...)$p.value; then do multiple testing correction with p.adjust(), then plot histogram of p-values and adjusted p-values; you can check [this code](https://github.com/microbiome/microbiome/blob/master/inst/extdata/check_wilcoxon.R) for some ideas

Let's first collect the data that we are going to use in analysis.
```{r}
# Does clr transformation. Pseudocount is added, because data contains zeros, and
# clr transformation includes log transformation.
tse <- transformCounts(tse, method = "clr", pseudocount = 1)

# Does transpose, so samples are in rows, then creates a data frame.
abundance_analysis_data <- data.frame(t(assay(tse, "clr")))

# Then we need variable for grouping samples. "Diet" column includes diet information.
# There two groups "western" and "control". Let's include that to the data frame.
abundance_analysis_data <- cbind(abundance_analysis_data, Groups = colData(tse)$Diet)

```

Next, let's check the normality of data. It can be done with Shapiro-Wilk test.
As we can see, data is not normally distributed.

I don't know if normality test is necessary.

```{r}
# Does Shapiro-Wilk test. Does it only for columns that contain abundances, not for
# column that contain Groups.

normality_test_p <- c()

for (column in 
     abundance_analysis_data[, !names(abundance_analysis_data) %in% "Groups"]){
  # Does Shapiro-Wilk test
  result <- shapiro.test(column)
  
  # Stores p-value to vector
  normality_test_p <- c(normality_test_p, result$p.value)
}

print(paste0("P-values over 0.05: ", sum(normality_test_p>0.05), "/", sum(normality_test_p<=0.05)))
```
Now we can do Student's t-test.
```{r}
# Does t-test. Does it only for columns that contain abundances, not for
# column that contain Groups.

t_test_p <- c()

# abundance_analysis_data_sub <- abundance_analysis_data[abundance_analysis_data$Groups == "KO_t1_western" |
#                                                          abundance_analysis_data$Groups == "WT_t2_western",]

for (column in 
     names(abundance_analysis_data[, !names(abundance_analysis_data) %in% "Groups"])){
  #print(column)
  result <- t.test(abundance_analysis_data[,column] ~ Groups, data = abundance_analysis_data)
   
   # Stores p-value to vector
   t_test_p  <- c(t_test_p , result$p.value)
}

t_test_p <- data.frame(taxa = 
                         names(abundance_analysis_data[, !names(abundance_analysis_data) 
                                                       %in% "Groups"]),
                       p_raw = t_test_p)

```

Because of multiple tests were mad, we need to adjust p-values. That is because, 
when more tests are done it is more probable that inference occur and significance
levels are higher than they actually are. By adjustment, the significance levels of
individual tests are stricter. Here we use Holm method.

```{r}
# Does the adjustment
t_test_p$p_holm <- p.adjust(t_test_p$p_raw, method = "holm")
```

```{r}

df = data.frame(x = c(t_test_p$p_raw, t_test_p$p_holm), 
               type=rep(c("raw", "holm"), c(length(t_test_p$p_raw), length(t_test_p$p_holm))))

t_test_plot <- ggplot(d) + geom_histogram(aes(x=x, colour=type)) +
  xlab("p-value")

t_test_plot
```



- Then add, for the same data, [DESeq2](https://github.com/microbiome/tutorials/blob/master/deseq2.Rmd) example with our present data (exclude ANOVA from that example code)

```{r}



```


- Add scatterplot that compares the p-values from both methods





