---
title: "Import data"
output: html_notebook
---


This notebook imports _biom_ data to a _TreeSummarizedExperiment_ object. 

You are expected to have the necessary data files in a local subfolder
"data".


# Initialization

[Install](install.R) the necessary R packages if you have not already
done it.

Then load the R packages.


```{r}
library("mia")
```

Define source file paths.

```{r}
biom_file_path <- "data/table-with-taxonomy.biom"
sample_meta_file_path <- "data/WesternDietSert_extendedTimeComparisonsv4.txt"
tree_file_path <- "data/tree.nwk"
```


Create TSE object from biom file and investigate it.

```{r}
tse <- loadFromBiom(biom_file_path)
```


# Investigate the R data object

We have now imported the data in R. Let us investigate its contents.

```{r}
print(tse)
```


Counts include the abundance table from biom file. The head() command
shows just the beginning of the data table for an overview.

```{r}
head(assays(tse)$counts)
```

rowdata includes taxonomical information from biom file

```{r}
head(rowData(tse))
```


We notice that the imported biom file did not contain sample meta data
yet, so it includes empty data frame

```{r}
head(colData(tse))
```



Let us adds phylogenetic tree.

```{r}
# Read the tree 
tree <- ape::read.tree(tree_file_path)

# Adds tree to rowTree
rowTree(tse) <- tree

# Check
head(tse)
```

Now rowTree includes phylogenetic tree:

```{r}
head(rowTree(tse))
```

Add sample meta data file

```{r}
# We use this to check what type of data is it
read.table(sample_meta_file_path)

# It seems like a comma separated file and includes headers
# Let us read it:
sample_meta <- read.table(sample_meta_file_path, sep = ",", header = TRUE)

# Convert it from data.frame to DataFrame
# (required for our purposes)
sample_meta <- DataFrame(sample_meta)

# Then it can be added to colData
colData(tse) <- sample_meta
```

Now colData includes sample metadata	

```{r}
colData(tse)
```


