---
title: "Import data"
output: html_notebook
---

This notebook imports _biom_ data to a _TreeSummarizedExperiment_ object. 

Data is from article Tengeler AC, Dam SA, Wiesmann M, Naaijen J, van Bodegom M, 
Belzer C, Dederen PJ, Verweij V, Franke B, Kozicz T, Vasquez AA & Kiliaan AJ (2020)
[**Gut microbiota from persons with attention-deficit/hyperactivity disorder affects the brain in mice**](https://doi.org/10.1186/s40168-020-00816-x)
Microbiome 8:44. 

Data is located in a "data" subfolder. Data consists of 3 files: 

- biom file that contains abundance table and taxonomy information
- csv file that contains sample metadata
- tre file that contains a phylogenetic tree.

## Initialization

[Install](install.R) the necessary R packages if you have not already
done it.

Then load the R packages.

```{r, results='hide'}
library("mia")
```

Define source file paths.

```{r}
biom_file_path <- "data/Aggregated_humanization2.biom"
sample_meta_file_path <- "data/Mapping_file_ADHD_aggregated.csv"
tree_file_path <- "data/Data_humanization_phylo_aggregation.tre"
```  


Create SE object from biom file and investigate it.

```{r}
se <- loadFromBiom(biom_file_path)
```  


## Investigate the R data object

We have now imported the data in R. Let us investigate its contents.

```{r}
print(se)
```


Counts include the abundance table from biom file. Let us just use first cols and rows.

```{r}
assays(se)$counts[1:3, 1:3]
```

Now rowdata includes taxonomical information from biom file. The head() command
shows just the beginning of the data table for an overview.

```{r}
knitr::kable(head(rowData(se)))
```

Taxonomic ranks are not real rank names. Let's replace those taxonomic classes 
with real rank names. 

In addition to that, taxa names include, e.g., '"k__' before the name, so let's
make them cleaner by removing them. 

```{r}
names(rowData(se)) <- c("Kingdom", "Phylum", "Class", "Order", 
                        "Family", "Genus")

# Goes through whole DataFrame. Removes '.*[kpcofg]__' from strings, where [kpcofg] 
# is any character from listed ones, and .* any character.
rowdata_modified <- BiocParallel::bplapply(rowData(se), 
                                           FUN = stringr::str_remove, 
                                           pattern = '.*[kpcofg]__')

# Genus level has additional '\"', so let's delete that also
rowdata_modified <- BiocParallel::bplapply(rowdata_modified, 
                                           FUN = stringr::str_remove, 
                                           pattern = '\"')

# rowdata_modified is list, so it is converted back to DataFrame. 
rowdata_modified <- DataFrame(rowdata_modified)

# And then assigned back to the SE object
rowData(se) <- rowdata_modified

knitr::kable(head(rowData(se)))

```

We notice that the imported biom file did not contain sample meta data
yet, so it includes empty data frame

```{r}
head(colData(se))
```

## Add side information

Let us add sample meta data file

```{r}
# We use this to check what type of data is it
# read.table(sample_meta_file_path)

# It seems like a comma separated file and includes headers
# Let us read it and then convert from data.frame to DataFrame
# (required for our purposes)
sample_meta <- 
sample_meta <- DataFrame(read.table(sample_meta_file_path, sep = ",", header = FALSE))
# Add sample names to rownames
rownames(sample_meta) <- sample_meta[,1]
# Delete column that included sample names
sample_meta[,1] <- NULL

# We can add titles 
colnames(sample_meta) <- c("patient_status", "cohort", "patient_status_vs_cohort", "sample_name")

# Then it can be added to colData
colData(se) <- sample_meta

```

Now colData includes sample metadata. Use kable to print it more nicely.

```{r}
knitr::kable(head(colData(se)))
```

Now, let's add a phylogenetic tree.

```{r}
# Current object, se, is a SummarizedExperiment object, thus, it does not include
# slot for phylogenetic tree. However, we can convert SE object to 
# TreeSummarizedExperiment which includes rowTree slot.
tse <- as(se, "TreeSummarizedExperiment")

# tse includes same data as se
print(tse)
```

```{r}
# Reads the tree 
tree <- ape::read.tree(tree_file_path)

# Adds tree to rowTree
rowTree(tse) <- tree

# Check
head(tse)
```

Now rowTree includes phylogenetic tree:

```{r, eval=FALSE}
head(rowTree(tse))
```