---
title: "Install and load required R packages"
output: md_document
---

# Install and load required R packages

In this section, all required packages are installed and loaded into the session.
If packages are already installed, installation step is skipped; only uninstalled
packages are installed. 

```{r warning = FALSE, message = FALSE}
# List of packages that we need from cran and bioc 
cran_pkg <- c("BiocManager", "dplyr", "ecodist", "ggplot2", "gridExtra", "knitr", "vegan")
bioc_pkg <- c("ANCOMBC", "ape", "DESeq2",  "DirichletMultinomial", "mia", "miaViz")

# Gets those packages that are already installed
cran_pkg_already_installed <- cran_pkg[ cran_pkg %in% installed.packages() ]
bioc_pkg_already_installed <- bioc_pkg[ bioc_pkg %in% installed.packages() ]

# Gets those packages that need to be installed
cran_pkg_to_be_installed <- setdiff(cran_pkg, cran_pkg_already_installed)
bioc_pkg_to_be_installed <- setdiff(bioc_pkg, bioc_pkg_already_installed)
```

```{r warning = FALSE, message = FALSE}
# If there are packages that need to be installed, installs them from CRAN
if( length(cran_pkg_to_be_installed) ) {
   install.packages(cran_pkg_to_be_installed)
}
```

```{r warning = FALSE, message = FALSE}
# If there are packages that need to be installed, installs them from Bioconductor
if( length(bioc_pkg_to_be_installed) ) {
   BiocManager::install(bioc_pkg_to_be_installed, ask = F)
}
```
 
Now all required packages are installed, so let's load them into session.
Some function names occur in multiple packages. That is why miaverse's packages
mia and miaViz are prioritized. Packages that are loaded before have higher priority.

```{r warning = FALSE, message = FALSE}
# Reorders bioc packages, so that mia and miaViz are first
bioc_pkg <- c(bioc_pkg[ bioc_pkg %in% c("mia", "miaViz") ], bioc_pkg[ !bioc_pkg %in% c("mia", "miaViz") ] ) 

# Loading all packages into session. Returns true if package was successfully loaded.
sapply(c(bioc_pkg, cran_pkg), require, character.only = TRUE)
```
 